---
title: "Lab #7: Modeling With Statistics"
author: "Scott Burstein"
date: "10/10/2020"
###"Data Science and Society (Sociology 367) - Lab #7"
output: html_document
---

In this lab, we will practice making statistical models of data using R. Throughout the lab we’ll be using these models to understand how Spotify calculates their “popularity” scores assigned to each artist. This is a synthetic score that spotify assigns to each song based on some features of user data. We start on the premise that Spotify may consider some of the features in our dataset when creating the score.

Edit the Lab #7 markdown file (link to file) for your submission. Remember to change the “author” field above to your name. Send the markdown file (.Rmd) to your TA via direct message on Slack. Be sure to do the required reading!

Required reading:

R for Data Science: Modeling (Chapters 23-25)
Optional reading:

Required Datasets:
We will use the following datasets for the exercises.

artist_meta:` table of Grammy nominee names and some of their spotify metadata.
Load the datasets and libraries. NOTE! It might be a good idea to convert the popularity column to a “double” type here to make later analyses easier. Do that here using the mutate and as.integer functions.

```{r load-data}
library(tidyverse)
library(MASS)
load(url('https://dssoc.github.io/datasets/artist_meta.Rdata'))
artist_meta <- artist_meta %>% mutate(popularity=as.integer(popularity))
head(artist_meta)
view(artist_meta)
```


## # A tibble: 6 x 6
##   artist   wikipedia_link       spotifyID   followers popularity genre          
##   <chr>    <chr>                <chr>           <int>      <int> <chr>          
## 1 Bon Iver https://en.wikipedi~ 4LEiUm1SRb~   2546477         79 chamber pop,ea~
## 2 Billie ~ https://en.wikipedi~ 6qqNVTkY8u~  20409690         99 electropop,pop 
## 3 Ariana ~ https://en.wikipedi~ 66CXWjxzNU~  42023952         95 dance pop,pop,~
## 4 H.E.R.   https://en.wikipedi~ 3Y7RZ31TRP~   2150235         83 alternative r&~
## 5 Khalid   https://en.wikipedi~ 6LuN9FCkKO~  10408465         96 alternative r&~
## 6 Lil Nas~ https://en.wikipedi~ 7jVv8c5Fj3~   2456187         86 country rap,lg~

***
#Exercises


1. In your own words, describe what statistical modeling means. When is it used? What does it allow data scientists to do?

**What are Statistical Models?**

Statistical models provide human interpretable information about data. Models are best when they provide insight about trends and patterns within the data, yet ignore data elements that are either erroneous or significantly distant from expected. This allows for a better understanding of the summary metrics by which the data can be best evaluated on.

Additionally, models can be either "predictive" or for "data discovery". In other terms these two types are defined as "supervised" and "unsupervised". Essentially, "predictive" or "supervised" models are used to make predictions about values that may not be present in the data set. "data discovery" or "supervised" are used to better understand the data given and to find patterns and trends within the set.


2. What is the correlation and statistical significance (measured by p-value) between an artist’s popularity and their number of followers on Spotify? Why would this correlation be less than 1.0?

Note: be sure you have converted the popularity column to numeric, as described above.

Hint: you may find the cor.test function helpful.

Prior to conducting a p-test, it is necessary to establish null and alternative 
hypotheses in order to understand what the p-value is actually revealing about 
the data.

$H_o:$ Artist's popularity and number of followers on Spotify are independent.

$H_a:$ Artist's popularity and number of followers on Spotify are not independent.

```{r pop-num-follow-corr}
cor.test(artist_meta$popularity, artist_meta$followers)

#correlation and statistical significance (p-value) are both in Pearson's corr.
```

The correlation between an artist's popularity and their number of followers on 
Spotify is 0.658. This makes sense since the more popular 
an artist is on Spotify, the more people will likely follow them. In fact, I am 
somewhat surprised the correlation is not higher. This reveals that there are 
other factors, perhaps like Social Media presence or popularity on other 
platforms, that generate more Spotify followers for artists who may not actually 
have as many active listeners on Spotify.

The p-value is <2.2e-16, which is effectively 0. Typically, a p-value less than 
0.05 is deemed statistically significant, as it is sufficient to reject the null 
hypothesis under this condition. It is important to note that this is NOT 
saying that the alternative hypothesis is true. Rather, there is only enough 
evidence to reject the null hypothesis. Hence, there is sufficient evidence to 
reject the notion that Artist's popularity and number of followers on Spotify 
are independent variables.

3. Create a scatter plot showing the number of followers on the x-axis and artist popularity on the y-axis. Display a trend line on the plot. Can you draw any rough quantitative information about the relationship between these two variables just from this plot?


```{r scatter-plot}
pop_plot <- ggplot(data = artist_meta,
                   aes(x = followers, y = popularity)) +
  geom_point() + 
  geom_smooth() +
  theme_minimal() +
  labs(title = "Artist Popularity vs. Followers Count on Spotify")

pop_plot

#When method and formula left empty,
#`geom_smooth()` using method = 'loess' and formula 'y ~ x'
```

It is challenging to draw explicit conclusions about the quantitative relationship 
between the `followers` and `popularity` variables. However, there is a 
relationship between the two, such that the number of followers has a small 
effect on popularity *up to a certain point*, when approximately 
followers = ~200,0000. After this threshold, artists who have more than this 
number of followers experience a rapid growth, acrruing a much larger audience 
base on the platform.

Another way to interpret this would be that an artist's followers count is 
relatively stagnant, until they reach a certain popularity around ~80, at which 
point they have "made it" and begin to generate followers much faster than 
before.

4. Generate a new column of logical type called is_pop to indicate whether a given artist is categorized in any pop genre (e.g. any genre that includes the word “pop”, like “rap pop”, etc). Create another column named is_rap that indicates whether the artist has been included in any “rap” genre. Make summary statistics indicating how many artists fall into either category.

**Instantiate is_pop variable:**
```{r is-pop-var}
artist_meta <- artist_meta %>%
  mutate(is_pop = grepl('pop', genre, fixed=TRUE)) %>%
  mutate(is_pop = ifelse(is_pop, 'pop', 'not pop')) %>%
  mutate(is_pop = as.factor(is_pop)) 
```

**Instantiate is_rap variable:**
```{r is-rap-var}
artist_meta <- artist_meta %>%
  mutate(is_rap = grepl('rap', genre, fixed=TRUE)) %>%
  mutate(is_rap = ifelse(is_rap, 'rap', 'not rap')) %>%
  mutate(is_rap = as.factor(is_rap)) 
```

**View new columns:**
```{r view-meta}
artist_meta
```

**Pop Count:**
```{r is-pop-count}
pop_count <- artist_meta %>%
  count(is_pop)
pop_count
```

**Rap Count:**
```{r is-rap-count}
rap_count <- artist_meta %>%
  count(is_rap)
rap_count
```

Out of the 141 artists, 97 are pop artists while 44 are not pop artists. 110 are 
rap artists while 31 are not rap artists. Each of the 141 artists must identify 
as either pop or not pop **and** either rap or not rap. These two are not 
mutually exclusive, but rather potentially independent. The relationship will be 
computed in exercise 5.

5. First compute a measure of the relationship (you may use correlations or chi-squared tests here) between the is_pop and is_rap variables. Then compute the relationships with each of those two variables and Spotify’s popularity metric. Which relationships are statistically significant? What does that tell us about Spotify’s measure of popularity?

Note: you should end up with three models - one for each pairwise similarity test.

### Chi-squared test for independence:
$$\chi^2 = \sum \frac {(O - E)^2}{E}$$

Chi-squared value reveals the extent to which two data sets are unique. A low 
value indicates strong correlation between the two sets. 

Conversely, a high chi-squared value indicates minimal correlation between the 
two sets.

The degrees of freedom (df) refers to the number of independent variables which 
affect the variables of interest in a given data set.

### For All Three Models:
#### (pop - rap | pop - popularity | rap - popularity)

$H_o:$ The two variables are independent.

$H_a:$ The two variables are not independent.

```{r pop-rap-chisq-test}
chisq.test(artist_meta$is_pop, artist_meta$is_rap)
```

This is a low X-squared value (0.91013), meaning there is some correlation 
between `is_pop` and `is_rap`. The p-value (0.3401) is too large though to reject 
the null hypothesis of no correlation.

Hence, it is improbable that a correlation exists between `pop` and `popularity` 
of a song.

```{r pop-popularity-chisq-test}
chisq.test(artist_meta$is_pop, artist_meta$popularity)
```

This is a high X-squared value (45.852), meaning there is not a strong correlation 
between `is_pop` and `popularity`. The p-value (0.316) is too large to reject 
the null hypothesis of no correlation.

Hence, it is improbable that a correlation exists between `pop` and `popularity` 
of a song.

```{r rap-popularity-chisq-test}
chisq.test(artist_meta$is_rap, artist_meta$popularity)
```

This is a high X-squared value (42.955), meaning there is not a strong correlation 
between `is_rap` and `popularity`. The p-value (0.4301) is too large to reject 
the null hypothesis of no correlation.

Hence, it is improbable that a correlation exists between `rap` and `popularity` 
of a song.

### Important Notes for Q5:
***
Was unable to produce cor() or cor.test() products because could not convert 
`is_pop` or `is_rap` to numeric type. Error was the following:
"'x' must be a numeric vector" - where x was the `is_pop` or `is_rap` count.

Also, the three chi-squared tests all yielded 
"Chi-squared approximation may be incorrect" warnings, which are concerning yet 
not problematic for the scope of this analysis.

Furthermore, I find it unlikely that all three of these models yielded no 
correlation. Perhaps the chi-squared approximation errors were in fact valid, 
and my analysis is flawed. There **should** be a correlation between these 
variables, as certain genres of music, such as pop and rap, are inherently more 
popular than others.

***

6. Use ggplot to create box plots or violin plots showing the popularity by is_pop and is_rap. What can you say about the data from these plots?

```{r pop-violin-plot}
pop_violin_plot <- ggplot(data = artist_meta,
                      aes(x = is_pop, y = popularity, fill = is_pop)) +
  geom_violin() +
  labs(title = "Spotify Artist Popularity by Pop Classification",
       x = "Artist Is Pop?",
       y = "Popularity (40-100)")
pop_violin_plot
```

Based on this violin plot, it is evident that artists classified as `pop` have 
higher Popularity on average than `not pop` artists. 

```{r rap-violin-plot}
rap_violin_plot <- ggplot(data = artist_meta,
                      aes(x = is_rap, y = popularity, fill = is_rap)) +
  geom_violin() +
  labs(title = "Spotify Artist Popularity by Rap Classification",
       x = "Artist Is Rap?",
       y = "Popularity (40-100)")
rap_violin_plot
```

Based on this violin plot, it is evident that artists classified as `rap` have 
higher Popularity on average than `not rap` artists. 

Compared to the *Spotify Artist Popularity by Pop Classification*, the 
*Spotify Artist Popularity by Rap Classification* reveals that the `not rap` 
artists have a tighter distribution around a Popularity mean of 80, whereas the 
`not pop` distribution in the former plot was more spread out. This potentially 
indicates that the artists deemed as `not rap` are more likely to also be `pop` 
or other genres that are also popular.

On the other hand though, the `rap` violin plot has a tight distribution around 
87, indicating that most rap artists are popular, even moreso than `pop` artists.

7. Create a multivariate linear model (see the lm function) predicting the Spotify popularity measure from both is_pop and is_rap. Then create a second model that additionally includes the artist’s number of followers as a covariate. What do these models tell us? Why would we create a second model that includes the additional variable?

***

### Multivariate linear model predicting Spotify popularity measure from `is_pop`:
```{r pop-popularity-lm}
summary(lm(popularity~is_pop,
           data = artist_meta))
```

### Multivariate linear model predicting Spotify popularity measure from `is_rap`:
```{r rap-popularity-lm}
summary(lm(popularity ~ is_rap,
           data = artist_meta))
```

***
## Adding number of followers as a covariate:
***

### Multivariate linear model predicting Spotify popularity measure from `is_pop` & `followers`:
```{r pop-followers-popularity-lm}
summary(lm(popularity ~ is_pop + followers,
           data = artist_meta))
```

### Multivariate linear model predicting Spotify popularity measure from `is_rap` & `followers`:
```{r rap-followers-popularity-lm}
summary(lm(popularity ~ is_rap + followers,
           data = artist_meta))
```
***

##Alternative Interpretation to Q7, modeling popularity based off of `is_pop` and `is_rap` covariates:
***

### Multivariate linear model predicting Spotify popularity measure from `is_pop` & `is_rap`:
```{r pop-rap-popularity-lm}
summary(lm(popularity ~ is_pop + is_rap,
           data = artist_meta))
```

### Multivariate linear model predicting Spotify popularity measure from `is_pop`, `is_rap` & `followers` variables:
```{r pop-rap-followers-popularity-lm}
summary(lm(popularity ~ is_pop + is_rap + followers,
           data = artist_meta))
```

It is challenging to decide explicitly which of the linear models are most 
convincing. I believe all of these covariates play a role in determining Spotify 
artist popularity, as all six linear models tested yield very small p-values, 
significantly less than 0.05, which means that for all of these variables, we can 
reject the null hypothesis that there is no correlation between any of the tested 
variables (`is_pop`, `is_rap`, `followers`) and the resultant `popularity` 
variable. Hence, it is likely that all of these affect the Spotify popularity of 
an artist in some way.


8. Name two unobserved but possibly important variables that we could not include in our model because we don’t have them in our dataset. Do any of our observed variables act as confounders for the unobserved variables? As in, might there be other features that Spotify uses to generate the popularity measure that we cannot see here, but nevertheless are represented in the model indirectly?

Hint: you may need to do research about unobserved and confounding variables to answer this question.

Note: these types of questions are critical to consider when assessing our data science models. Be sure to think about them when you examine the results of your data science project.

Spotify's popularity index likely has many other variables that are not present 
in the `artist_meta` table. For example, an artist who releases an album and has 
an uptick in the number of listeners will likely increase their popularity index. 
If certain songs by the artist garner a lot of attention and listens on Spotify, then 
they will likely become more popular. All of these variables are likely confounding, 
as the more listeners an artist has, it is likely the more followers they will accrue and their 
popularity rating will also increase. 

Another variable is presence on a number of Spotify's 
generated playlists and/or top charts. When an artist is featured on these platforms, they will 
get more listeners and followers, which will also increase popularity.

Both an artist's release record and their presence on Spotify's popular playlists and charts will impact popularity. These variables definitely confound with the number of followers an artist has, since these measures are all correlated with one another. When an artist is doing well and releasing new music, they will have more listeners, followers, and popularity.

9. How could you use statistical modeling to answer the hypothesis you provided in the last assignment? What inferences could you make?

I could use statistical modeling to answer my hypothesis in the previous question 
by establishing a null (there is no correlation) and an alternative (there is correlation) 
hypotheses for each of the covariate relationships with overall `popularity`. Then, 
I could conduct any number of tests, including the p-test and chi-squared test to determine 
p-value and correlation between the different variables.

By doing so, I could infer how correlated these different conditions are for an artist 
and their respective popularity.

10. Describe one or two existing datasets that you would like to use for the project you’ve been developing last week. Will you be able to download the data from somewhere, or can you use an API? Will you be making statistical models, analyzing networks, doing text analysis, or creating visualizations? See the “Final Project” section in the course description page on the website.

https://www.kaggle.com/headsortails/us-natural-disaster-declarations

This is an interesting dataset from Kaggle that has al U.S. natural disaster 
declarations, date/time, and other details starting in 1953.

https://cran.r-project.org/web/packages/gtrendsR/gtrendsR.pdf

I could combine this with Google Trends (valid for recent years) and the 
gtrendsR API package, which is the CRAN package for using Google Trends in R.
https://trends.google.com/trends/?geo=US

Combining these two sources (natural disaster data set and Google Trends insights) 
could yield interesting analysis of recent shifts in human behavior, such as search patterns 
surrounding natural disasters.

Geographic visualizations would be very applicable, such as laying out recent natural disasters on a map. I could then associate this with multiple statistical models to see if there were shifts 
in Google search trends surrounding these events.

*Potential questions could include:*

-During a hurricane, what to people Google search?

-Are certain events more likely to create demand for certain products?

-Are certain events or regions more prone to learn about climate change after 
natural disasters occur?

It would also be interesting to quantify overall patterns in search trends during these events, 
such as how the quantity of search querries on Google changes during certain events.

