---
title: "Grammy Awards Programming Lab"
author: "Scott Burstein"
date: "09/19/2020"
###"Data Science and Society (Sociology 367) - Lab #4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this lab, we will practice the fundamentals of programming. Edit the Lab #4 markdown file ([link to file](https://github.com/dssoc/dssoc.github.io/raw/master/assignments/Lab_4.Rmd)) for your submission. Remember to change the "author" field above to your name. Send the markdown file (.Rmd) to your TA via direct message on Slack. Be sure to do the required reading!

**Required reading:** [R for Data Science: Basic Programming (17 -21)](https://r4ds.had.co.nz/program-intro.html).

**Optional reading:** [More on functions](https://rstudio.cloud/learn/primers/6).


### Required Datasets

The tables that you will be drawing on for this lab can be downloaded from the course website: 

(1) `nominees`: [grammy_nominees_20102020.Rdata](https://github.com/dssoc/dssoc.github.io/raw/master/datasets/grammy_nominees_20102020.Rdata) is a table of Grammy nominees from 2010 to 2020 in the categories "Album of the year" (album_oty), "Record of the year" (record_oty), "Song of the year (song_oty), and "Best new artist of the year" (new_artist_oty).
(2) `winners`: [grammy_winners_20102020.Rdata](https://github.com/dssoc/dssoc.github.io/raw/master/datasets/grammy_winners_20102020.Rdata) is a table of Grammy winners for the above listed categories, from 2010 to 2020.


The exercises in this lab correspond with R for Data Science, [chapters 17 - 21](https://r4ds.had.co.nz/program-intro.html):


Load the libraries and datasets:
```{r message=FALSE}
library(tidyverse)
load('~/Downloads/grammy_nominees_20102020.Rdata')
load('~/Downloads/grammy_winners_20102020.Rdata')
```
<br/><br/>

### Exercises
<br/>

**1. In your own words, describe what Boolean operators (also called Logical operators) are, and provide one example of how you might use them in a data science project.**
```
Boolean operators, or logical operators, are any of the following in R: &, !, |, isTRUE(), isFALSE(), &&, ||, or(x,y). Decoding what these symbols and combinations of symbols means is necessary to determine the net boolean value of the operation. These originate from the CRAN library documentation. Some of the less obvious are !, which references unary not. | references Or, specifically for binary vectorized comparisons, where as || references not vectorized comparisons. & is traditinally read as "and", and && rerefences and as well, however && is - like || - for vectorized comparisons. These boolean or logical operators work in tandem with actual boolean values (TRUE/FALSE) that can be made based on comparisons such as ==, >, <, >=, <=, !=, etc.

These operators can be used in a data science project to, for example, filter for only certain rows of a dataset that contain a certain element or equal a specific value. The operators allow for more complex Boolean structures to be evaluated, such as those with multiple components.

#CRAN manual documentation can be found here:
#https://cran.r-project.org/doc/manuals/r-release/R-lang.pdf
```
<br/>

**2. In your own words, describe what a function is, and provide one example of how you might use it in a data science project.**
```
Functions in the R programming language allow for more dynamic and intuitive methods for analyzing large amounts of data. For example, to access the value of a specific formula applied to all rows in a dataset, a function would allow for easier human and computer comprehension. As opposed to, for example, copying/pasting the same function with slightly different inputs for each row in the dataset.

Functions allow for a structured approach to performing the same operation repeatedly on different inputs. According to Hadley Wichkahm's R for Data Science resource, functions can make code easier to understand, more modifiable as needs change, and less prone to error.
#https://r4ds.had.co.nz/functions.html

Functions can be used in data analysis anytime the data scientist wants to view a specific output or value for multiple independent rows, columns, tables, etc. of data. It is advised to use a function to replace any code that would need to be iterated 3 or more times in a given data science project.
```
<br/>

**3. Most of the time, packages in R contain many useful functions/commands. If you didn't know what a certain function did, or how it worked, where within R Studio would you look to learn more / see example code? Where would you look outside RStudio?**
```
RStudio's built in Help panel in the lower right quadrant allows for fast access to CRAN and library Help resources for every built in and imported function and command. Help for a given function, for example, can be located by searching `?___` where `?` is followed the function or variable of interest. Using ?? brings you to a list of plausible help pages, whereas ? brings you to the desired one if RStudio is able to identify it.

Outside of RStudio, online resources like Stack Overflow are ideal, as people constantly post questions and answers which are vetted by the communities' voting/rating system.
```
<br/>

**4. Write a function that takes a character vector as an argument and returns a character vector containing the first letters of each element in the original vector. To show that it works, test it on the character vector `sentence` defined below.**
```{r first-char-func}
sentence <- c('you', 'only', 'understand', 'data', 'if', 'data', 'is', 'tidy')

#this is simplest, however does not require implementing a fxn, 
#since it is already built into the substring() syntax:

#substring(sentence, 1, 1 )

first_char_func <- function(str_vec) {
  substring(str_vec, 1, 1 )
}
first_char_func(sentence)

```
<br/>

Other proposed solution:
first_chars <- function(str_vec) {
  output <- vector("character", length(str_vec))
  for (i in length(str_vec)) {
    output[i] <- str_vec[i]
  }
  output
}
first_chars(sentence)
vector

^ Not returning correct value

I tried using this approach first in order to not use the built-in substring() method, but I encountered issues with iterating through the str_vec and getting a correct vector. When I tried running this code, I got 7 empty strings followed by "tidy", which leads me to believe the output vector is resetting/not properly storing all 8 string values, however I am unsure.

**5. In your reading, you learned about function arguments. Using the `winners` table, write a function that takes year (numeric) as an argument and returns the name of all winners for that year. It's alright if the function simply returns the row for that year in the `winners` table, but you can also print each winner name if you'd like.**
```{r winners-for-year}

winners_for_year <- function(input_year) {
  filter(winners, year == input_year)
}
winners_for_year(2020)
```
<br/>

**6. Sometimes we would like to execute different commands or calculations depending on the structure of our data. You learned about this in the *conditional execution* section of your reading. Expand your function from Exercise 5 to return "Date not available" if a year is supplied to the function but is not in the dataset.**
```{r expanded-winners-for-year}
winners_for_year <- function(input_year) {
  
  if (input_year %in% winners$year) {
    filter(winners, year == input_year)
  }
  
  else {
    print("Date not available")
  }
}

#functionality still retained for valid years
winners_for_year(2020)
#invalid years return "Date not available"
winners_for_year(2025)
```
<br/>


**7. Now, using the `nominees` table, write code to calculate the percent of nominations that have only one performer. For this, you might want to consider using the `is.na()` function that you learned about in your reading.**
```{r percent-one-performer}
perc_noms_w_1_performer <- nominees %>%
  count(is.na(performer2)) / nrow(nominees)
perc_noms_w_1_performer %>%
  select(n)

```
Based on the data, 84.24% of all nominations had one performer. Conversely, only 15.76% of nominations had more than one performer.

<br/>

**8. As in one of the previous labs, below I define two vectors corresponding to policies that US States have adopted to respond to COVID-19: restrictions on travel ([recorded May 20, 2020](https://www.wsj.com/articles/a-state-by-state-guide-to-coronavirus-lockdowns-11584749351)) and requirements that citizens to wear masks in public ([recorded August 17, 2020](https://www.cnn.com/2020/06/19/us/states-face-mask-coronavirus-trnd/index.html)). Run this code to create the vectors yourself.**
```{r defined-two-vectors}
travel_restrictions <- c("WA", "OR", "NV", "CA", "NM", "MN", "IL", "OH", "MI", "PA", "VA", "NY", "MA", "VH", "ME", "DE", "MD", "NJ")

require_masks <- c("HI", "WA", "OR", "NV", "CA", "MT", "CO", "NM", "KS", "TX", "MN", "AR", "LA", "WI", "IL", "AL", "MI", "IN", "OH", "KY", "WV", "NC", "VA", "DC", "DE", "PA", "NY", "VT", "NH", "MA", "RI", "CT", "ME")
```

**Iterate over the states which issued travel restrictions, and print "[state abbreviation] issued travel restrictions but didn't require masks" if they did not require masks.**
```{r iterate-over-states}
#only_travel_restrictions is same syntax as only_travel from original analysis lab.
only_travel_restrictions <-
  setdiff(travel_restrictions, require_masks)

#only_travel_restrictions


for (state in only_travel_restrictions) {
  print (paste(state, "issued travel restrictions but didn't require masks"))
}
```
<br/>

**9. Change one element or step in your answer to Exercise 8 to provide an alternative way of completing the exercise. For instance, if you did not use a for loop in your first attempt, perhaps use one now. If you did use a for loop in your first attempt, perhaps consider using `paste()` and `setdiff()` instead.**
```{r change-one-element}

`%notin%` <- Negate(`%in%`)
for (state in travel_restrictions) {
  if (state %notin% require_masks) {
    print (paste(state, "issued travel restrictions but didn't require masks"))
  }
}
```
<br/>


I used a for loop as well as paste() and setdiff() in my original attempt.
this new version instead of setdiff() has a nested if() statement witin the for() loop in order to select only those states which are in travel_restrictions but not within require_masks.
I created a  %notin% operator by assigning the value Negate(`%in%`) to a more intutive operator name `%notin%`.


**10. Perform a transformation of your choice on one of the datasets provided for this Lab. Annotate your code to describe what your transformation does to someone who might later read your code (e.g. the TA ;) As a reminder, you should be at least modestly annotating all of your Lab answers.**
```{r data-transformation-explained}
#winners_extended produced using gather() in same way as in original Grammy Award Analysis Lab (winners_long).
winners_extended <- winners %>%
  gather(key = 'award', value = 'artist', `record_oty` : `new_artist_oty`)
winners_extended

#song_artist_by_year returns all song_oty awards for each year where awards were recorded in the dataset (2010-2020).
song_artist_by_year <- winners_extended %>%
  filter(award == "song_oty")
song_artist_by_year
```
The transformation done was an extension of the winners_long transformation done in the original Grammy Award Analysis lab. I sued gather() on the winners table to make more rows with fewer columns. I then filtered just for song_oty awards and the resulting table is all artists who received song_oty awards for each year from 2010-2020.