---
title: "Coronavirus Tweets Data Analysis"
author: "Scott Burstein"
date: "08/29/2020"
output: html_document
---
### Data Science and Society (Sociology 367) - Lab #1
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<br>
To edit this markdown document (you can copy and paste the text form [this link](https://github.com/dssoc/dssoc.github.io/raw/master/assignments/Lab_1.Rmd) into a new `.Rmd` file or write a new one from scratch. 

To complete these exercises, you'll need to download the `Coronavirus_Tweets.Rdata` dataset from [this link.](https://github.com/dssoc/dssoc.github.io/raw/master/datasets/Coronavirus_Tweets.Rdata)

<br>
**1. Load the "Coronavirus_Tweets.Rdata" dataset. Code that calculates the mean value of the variable "retweet_count" in the code chunk below**

```{r}
load("Coronavirus_Tweets.Rdata")
mean(covid_tweets$retweet_count)
```
<br>
**2. Why does the code below create an error message? Re-write the code so that it does not create an error message.**

The code below created an error message because the My_Variable variable was misspelled as "My_Var1able", which does not exist. The correct variable call is below.

```{r, eval=FALSE}
My_Variable<-1
My_Variable
```
<br>
**3. Extract the third element of the vector.**

```{r}
Vector_Saurus_Rex<-c("R","is","great")
Vector_Saurus_Rex[3]
```
<br>
**4. Write code that identifies the largest number in this vector (hint: you'll have to research some new functions in r to do this).**

```{r}
Large_Vector<-c(25555,342343,123123123,4234234,53243234,54324234,5421111,12312312,111231,
                     1231231,12312312,12312312,123123,898972,789872,2343,23423423,2343221,23423,
                     14444,44324222,2342341,124231111,22233345,1111233333,1231231,1231231)

max(Large_Vector)
```
<br>
**5. Calculate the square root of 2456.**
```{r}
sqrt(2456)
```
<br>
**6. Which Tweet in the Coronavirus Tweets.Rdata dataset received the most "likes" (or "favorites")? Code below identifies the tweet and displays the text of the tweet.**
```{r}
sorted_tweets <- covid_tweets[order(covid_tweets$favorite_count, decreasing=TRUE),]
pop_tweet <- (sorted_tweets$tweet_text)[1]
pop_tweet
num_favorites <- max(covid_tweets$favorite_count)
num_favorites
```
<br>
<br>
<br>